\begin{abstract}
At test time, an LLM theorem prover can spend the same remaining budget on
another whole-proof sample, a local repair, retrieval, decomposition, or a
stronger model. The control problem is therefore not only how long to search,
but which verifier-mediated computation to purchase next. We cast Lean~4 proof
search as budget-constrained metareasoning under explicit limits on model
requests and tokens, checker calls, latency, and monetary cost. We introduce
\system, a controller that schedules heterogeneous proof actions using
estimates of marginal solve value and incremental cost. The estimates are
grounded in a replayable workspace that records obligation dependencies,
alternative branches, checker observations, and versioned accepted artifacts.

In an idealized independent-candidate model, selecting candidates under a hard
resource budget is a multidimensional knapsack problem. Conditional on a
selected feasible set and a scalar accounting cost, decreasing \(p_i/c_i\)
minimizes expected expenditure until success or exhaustion. If an expensive
fallback of cost \(C\) is reserved and remains feasible, a cheap action of cost
\(c\) and success probability \(p\) reduces expected expenditure exactly when
\(c<pC\). These scoped results motivate, rather than establish the global
optimality of, a controller based on immediate acceptance, residual verifier
progress, and downstream unlock value.

We evaluate \system{} under matched resource vectors against independent
sampling and a strong iterative-refinement agent. Controlled comparisons hold
the structured substrate fixed while separately varying the action mask,
cost-aware selector, remaining-budget admission, and cheap/strong model router.
\todo{State absolute solve rates, confidence intervals, the matched budget,
and the task regime in which allocation gains exceed controller overhead.}
\end{abstract}
