\section{Introduction}

Formal theorem proving is an unusually clean setting for studying test-time
control. A language model may propose an invalid proof, but a proof assistant
such as Lean~4 can reject it with exact diagnostics or certify it with a small
trusted kernel \cite{demoura2021lean4}. Repeated computation can therefore be
useful in several different ways: it can generate an independent candidate,
repair a type error, retrieve an exact declaration, split a theorem into
helpers, or revise a mathematical representation. Each choice consumes
resources, and each can change the value of later choices.

A fixed test-time budget does not determine which of these computations an
agent will perform. With the same token and checker-call limits, one policy may
draw independent whole-proof samples, another may repeatedly refine one
candidate, and a third may reserve expensive calls for difficult subgoals.
Recent work makes this distinction consequential. A minimal agent based on
iterative refinement, compact context, and library search is already a strong
baseline \cite{requena2026minimal}; a closely related cost-aware prover shows
that failed Lean trajectories can guide continue-versus-redecompose decisions
\cite{rognvaldsson2026costquality}. More broadly, adaptive allocation can make
test-time scaling substantially more efficient than a fixed sampling schedule
\cite{snell2025scaling}. These results sharpen the question studied here:

\begin{quote}
\emph{Given the current verifier-backed state and a remaining resource vector,
which proof-search computation should be executed next?}
\end{quote}

We treat this as a metareasoning problem: proof-search computations are
themselves actions, their outcomes are uncertain, and the controller chooses
among them before the resource vector is exhausted. The objective is not to
minimize cost in isolation---which would reward doing nothing---but to maximize
the probability of returning a checker- and safety-accepted proof under fixed
hard limits. Cost becomes a secondary objective at matched success, or the
horizontal axis of a solve--cost frontier.

Our central claim is deliberately empirical and conditional:

\begin{quote}
\emph{Checker-grounded allocation can improve pass under a matched resource
budget only when the value of better action choices exceeds the full cost of
maintaining, prompting, and checking the controller's state.}
\end{quote}

Testing this claim requires separating three mechanisms that are often bundled
together. An \emph{action space} determines whether the agent may repair,
retrieve, decompose, or escalate at all. A \emph{selector} decides which
available computation receives the next unit of budget. A \emph{router}
chooses the model tier used for that computation. A richer system can beat a
baseline because it has better actions or a stronger model even when its
allocation rule contributes nothing. Our evaluation therefore varies these
mechanisms in controlled comparisons under the same structured substrate and
resource limits.

We instantiate the formulation in \system. The controller allocates model
requests, tokens, checker calls, retrieval, wall time, and monetary cost across
proof implementation, local repair, capability testing, decomposition,
argument refinement, representation change, and model escalation. It enforces
each hard resource limit directly and preserves a reserve for final integrated
checking. Within the feasible action set, it uses frozen cost estimates and
verifier-grounded signals to prioritize computations.

Those signals come from a structured proof workspace. Obligations form AND
dependencies, alternative arguments or implementations form OR branches, and
accepted artifacts are versioned and linked to checker evidence. The workspace
also records mathematical argument steps, Lean artifacts, alignment links,
diagnostics, and competing failure hypotheses. This structure is not the claim
by itself; it is instrumentation that exposes readiness, provenance, stalled
goals, downstream unlocks, and residual-progress features that a flat attempt
history would otherwise have to reconstruct. The language model remains a
single proof agent so that mathematical reasoning and Lean repair share the
same projected context.

Simple special cases provide useful checks on the controller without implying
global optimality. Under independent, state-invariant candidates, selecting a
feasible subset under a hard resource budget is a multidimensional knapsack
problem. Once that subset is fixed, decreasing \(p_i/c_i\) minimizes expected
scalar expenditure before success or exhaustion. Similarly, a cheap action
should precede an expensive fallback only when the fallback remains feasible
and the cheap action's expected avoided fallback cost exceeds its own cost.
Real proof actions violate the static assumptions: failures reveal information,
repairs alter later success probabilities, and decomposition changes the action
space. We therefore use residual verifier progress and unlock value as
empirical surrogates and test their calibration rather than presenting the
resulting heuristic as a globally optimal policy.

Our contributions are:

\begin{itemize}
    \item We cast LLM-guided Lean proof search as a multi-resource,
    budget-constrained metareasoning problem. We distinguish hard-budget
    candidate selection from within-set ordering and characterize scoped
    \(p_i/c_i\) and cheap-first rules.
    \item We instantiate this formulation in a replayable controller over
    heterogeneous proof actions. Its decisions combine vector budget
    admission, frozen incremental-cost estimates, and features derived from
    obligation dependencies, alternatives, diagnostics, and accepted
    artifacts.
    \item We design a matched-budget evaluation against independent sampling
    and a strong iterative-refinement baseline. Controlled mechanism studies
    separately measure structured-state overhead, action-space expansion,
    cost-aware selection, remaining-budget admission, and model-tier routing.
    \todo{Insert the main absolute result and confidence interval.}
\end{itemize}
