\section{Related Work}

\paragraph{Resource-bounded metareasoning and test-time allocation.}
Rational metareasoning treats computation itself as an action whose value
depends on how it changes a later decision under limited resources
\cite{horvitz1987resource,russell1991metareasoning,hay2012selecting}.
This perspective also makes clear why globally optimal control is generally
intractable: the value of one computation depends on the sequence of
computations it enables. In current LLM systems, cascades and routers trade
model quality against inference cost \cite{chen2023frugalgpt,ong2024routellm},
while compute-optimal test-time scaling allocates different amounts of search
to inputs of different difficulty \cite{snell2025scaling}. We do not claim to
introduce metareasoning or model routing. We instantiate them in formal proof
search, where the controlled computations include verifier-mediated repair,
retrieval, decomposition, and model escalation, and where feasibility is
defined by a vector of hard resource limits.

\paragraph{Neural and LLM-guided theorem proving.}
Early generative provers demonstrated that language models can synthesize
machine-checkable proofs \cite{polu2020gptf}. In Lean, LeanDojo and ReProver
made retrieval-augmented proof search reproducible at scale
\cite{yang2023leandojo}, while Lean Copilot integrated learned suggestions,
goal completion, and premise selection into Lean workflows
\cite{song2024leancopilot}. Large-scale systems such as DeepSeek-Prover and its
successors improve proving through synthetic data, reinforcement learning,
subgoal decomposition, and high test-time sampling
\cite{xin2024deepseekprover,ren2025deepseekprov_v2}. Goedel-Prover and
Goedel-Prover-V2 similarly combine data synthesis and verifier-guided
self-correction \cite{lin2025goedel,lin2025goedelv2}. These works establish the
capability of the underlying provers. Our object of study is the control policy
that allocates a fixed inference budget among heterogeneous proof actions.

\paragraph{Verifier-guided agents and decomposition.}
Compiler feedback can make iterative proof generation more sample-efficient
than independent generation. APOLLO combines Lean feedback, proof repair,
subgoal isolation, and automated solvers under a limited sampling budget
\cite{ospanov2025apollo}. A Minimal Agent shows that iterative refinement,
compact context management, and library search form a strong baseline even
without an elaborate hierarchy \cite{requena2026minimal}. We use this result
as a baseline-strength requirement rather than a claim to refute: a structured
controller must justify its extra state and prompt cost under a matched budget.
Compile to Compress uses structured compiler failure modes for local refinement
and value-guided search \cite{li2026compile}, while Goedel-Architect constructs
and refines dependency blueprints of definitions and lemmas
\cite{chung2026architect}. These systems motivate verifier feedback and
explicit dependencies, but their action space, state representation, and
scheduling policy are evaluated together. Our mechanism study holds the
structured substrate fixed while varying which actions are exposed and how
they are selected.

\paragraph{Proof progress and cost-aware routing.}
HyperTree Proof Search uses learned search to improve neural theorem proving
\cite{lample2022hypertree}, and LeanProgress studies proof-progress prediction
as a search signal \cite{george2026leanprogress}. A closely related concurrent
approach estimates the success and cost of another proof attempt, then routes
between continuing the current target and restarting from a new decomposition
\cite{rognvaldsson2026costquality}. We study a broader control interface with
multiple hard resources, a typed set of local and structural actions, and an
explicit AND/OR workspace. The important empirical distinction is not the size
of this interface by itself: our ablations separately test expanded actions,
cost-aware selection, remaining-budget admission, and model-tier routing.

Outside theorem proving, Budget-Aware Value Tree Search conditions node
selection on remaining resources and uses residual rather than absolute value
estimation \cite{li2026bavt}. Budget-Guided MCTS changes its search behavior as
a fixed token budget is depleted \cite{miyamoto2026bgmcts}. Lean feedback adds
typed failure evidence, stable goal fingerprints, accepted intermediate facts,
and exact terminal correctness. These signals make marginal computation value
more observable, but they do not make a heuristic progress score equal to true
distance from a proof.

\paragraph{Benchmarks.}
miniF2F provides cross-system Olympiad-style formal statements
\cite{zheng2021minif2f}; ProofNet pairs natural-language mathematics with Lean
statements and proofs \cite{azerbayev2023proofnet}; and PutnamBench targets
substantially harder undergraduate mathematics
\cite{tsoukalas2024putnambench}. Our primary controller comparison fixes the
formal statement and imports so that autoformalization errors do not confound
resource allocation.
