\section{Budget-Constrained Proof Search}

\subsection{Metalevel Interaction Model}

Let \(W_t\) be the verifier-grounded proof workspace after \(t\)
computations, \(\mathcal H_t\) the action and cost history visible to the
controller, and \(\mathbf b_t\) the remaining resource vector. The metalevel
state is
\[
    s_t=(W_t,\mathcal H_t,\mathbf b_t).
\]
At each step the controller chooses \(a_t\in\mathcal A(s_t)\). Executing it
returns a joint random outcome--cost pair
\[
    (O_t,\mathbf C_t)\sim P(\cdot\mid s_t,a_t),
\]
after which
\[
    W_{t+1}=T(W_t,a_t,O_t),
    \qquad
    \mathbf b_{t+1}=\mathbf b_t-\mathbf C_t.
\]
The reducer \(T\) is deterministic given the action and observed tool result;
uncertainty enters through model generations, retrieval results, checker
outcomes, and realized token or wall-clock use.

The resource vector records
\[
  \mathbf C =
  (C_{\mathrm{request}},C_{\mathrm{in}},C_{\mathrm{out}},
   C_{\mathrm{billed}},C_{\mathrm{check}},C_{\mathrm{retrieve}},
   C_{\mathrm{time}},C_{\mathrm{money}}).
\]
Some coordinates are correlated or derived, but retaining them separately
prevents a low monetary price from concealing excessive checker calls, tokens,
or latency. Hard limits are enforced component-wise. Before execution, a
configured upper bound \(\overline{\mathbf c}(s,a)\) is used for admission and
must leave a reserve \(\mathbf b^{\mathrm{reserve}}\) for final assembly. The
estimate \(\hat{\mathbf c}(s,a)\) used for ranking is distinct from both this
admission bound and the realized \(\mathbf C_t\).

Our action space includes proof implementation, local repair, retrieval,
capability testing, decomposition, argument refinement, representation change,
model escalation, branch pruning, assembly, and stopping. Let
\(Y_\pi(x,\mathbf B)\in\{0,1\}\) indicate whether policy \(\pi\) produces a
proof accepted by Lean and by deterministic safety review before any component
of initial budget \(\mathbf B\) is exhausted. The primary objective is
\[
  \max_\pi\ \mathbb E_{x\sim\mathcal D}
  [Y_\pi(x,\mathbf B)].
\]
Equivalently, the ideal value-to-go is \(V^\star(W,\mathbf b)=1\) when
\(W\) contains an assembled checker- and safety-accepted proof. At a
nonterminal state it obeys the metalevel recursion
\[
\begin{aligned}
Q^\star(W,\mathbf b,a)
  &=\mathbb E\!\left[
      V^\star\!\left(T(W,a,O),\mathbf b-\mathbf C\right)\right],\\
V^\star(W,\mathbf b)
  &=\max\!\left\{0,
      \sup_{a\in\mathcal A_{\mathrm{feas}}(W,\mathbf b)}
      Q^\star(W,\mathbf b,a)\right\}.
\end{aligned}
\]
The transition law and value function are unknown and the state space is far
too large to solve this recursion exactly. \system{} is an auditable approximate
policy for this constrained metareasoning problem.

We evaluate the objective at preregistered, component-wise comparable budget
vectors. For efficiency claims, expected realized resource use per attempted
task is compared subject to a preregistered non-inferiority margin on accepted
rate. Resource use conditional on success and amortized resource use per
accepted proof are descriptive secondary quantities; they are not substituted
for the fixed-budget objective.

\subsection{Static Candidate Selection and Ordering}

A state-invariant special case separates two decisions that are easy to
conflate. Suppose candidate \(i\) has deterministic resource cost
\(\mathbf c_i\succeq\mathbf 0\), independently succeeds with probability
\(0\le p_i<1\), and neither its cost nor success probability depends on the
order in which candidates are evaluated.

\begin{proposition}[Candidate selection under a hard budget]
Selecting a feasible candidate subset to maximize the probability of at least
one success is equivalent to
\[
  \max_{S\subseteq\{1,\ldots,n\}}
  \sum_{i\in S}-\log(1-p_i)
  \quad\text{subject to}\quad
  \sum_{i\in S}\mathbf c_i\preceq\mathbf B.
\]
Thus even the static multi-resource selection problem is a multidimensional
\(0\)--\(1\) knapsack problem in general.
\end{proposition}

\begin{proof}
For a subset \(S\), independence gives success probability
\(1-\prod_{i\in S}(1-p_i)\). Maximizing this quantity is equivalent to
minimizing its failure probability and hence, by monotonicity of the logarithm,
to maximizing \(\sum_{i\in S}-\log(1-p_i)\) under the same constraints.
\end{proof}

Once a feasible subset is fixed, ordering controls expected realized
expenditure because later candidates are skipped after success. Let
\(c_i>0\) be a specified scalar accounting cost, such as one measured resource
or \(c_i=\boldsymbol\lambda^\top\mathbf c_i\) for a fixed nonnegative shadow
price vector \(\boldsymbol\lambda\).

\begin{theorem}[Optimal ordering of a fixed candidate set]
Assume each candidate is evaluated at most once, evaluation stops at the first
success or after all candidates fail, and the success probabilities of
unevaluated candidates are unchanged after any common failed prefix. Mutual
independence is sufficient. Every ordering \(\sigma\) that is nonincreasing in
\(p_i/c_i\) minimizes
\[
  \mathbb E[C_\sigma]
  =
  \sum_{k=1}^{n}c_{\sigma(k)}
  \prod_{\ell<k}(1-p_{\sigma(\ell)})
\]
among permutations of the same candidate set. Ties may be broken arbitrarily,
and zero-probability candidates may be placed last.
\end{theorem}

\begin{proof}
Condition on reaching two adjacent candidates \(i\) and \(j\). Their expected
additional costs are \(c_i+(1-p_i)c_j\) and
\(c_j+(1-p_j)c_i\), respectively. The former is no larger exactly when
\(p_i/c_i\ge p_j/c_j\). The probability of reaching the pair is a common
nonnegative factor, so repeated adjacent swaps remove every inversion.
\end{proof}

The theorem is an ordering result, not a claim that greedy selection by
\(p_i/c_i\) maximizes Pass@budget. In the scalar fractional relaxation, the
selection density is \(-\log(1-p_i)/c_i\), which approaches \(p_i/c_i\) only
for small \(p_i\). With multiple hard resources there is no canonical scalar
ratio at all. We therefore use \(p/c\) as a local policy target only after
component-wise feasibility filtering.

\subsection{A Feasible Cheap Prefix}

The same distinction clarifies when cheap search should precede escalation.
Let a cheap action \(L\) have deterministic resource cost \(\mathbf c_L\) and
immediate success probability \(p_L\). Conditional on its failure, suppose an
expensive action \(H\) remains available with deterministic cost
\(\mathbf C_H\).

\begin{proposition}[Expected-cost criterion for cheap-first execution]
Assume the worst-case sequence preserves the final reserve,
\[
  \mathbf c_L+\mathbf C_H
  \preceq
  \mathbf b-\mathbf b^{\mathrm{reserve}}.
\]
The policy \(LH\), which executes \(H\) only after \(L\) fails, has expected
resource use
\[
  \mathbb E[\mathbf C_{LH}]
  =\mathbf c_L+(1-p_L)\mathbf C_H.
\]
For any fixed nonnegative scalarization \(\boldsymbol\lambda\), \(LH\) has
lower expected scalarized cost than direct \(H\) if and only if
\[
  \underbrace{\boldsymbol\lambda^\top\mathbf c_L}_{c_L}
  <
  p_L\underbrace{\boldsymbol\lambda^\top\mathbf C_H}_{C_H}.
\]
It Pareto-improves expected resource use when
\(c_{L,j}\le p_L C_{H,j}\) for every resource \(j\), with strict inequality
for at least one component.
\end{proposition}

This is an expected-resource criterion, not by itself a hard-budget success
criterion. If \(L\) and \(H\) do not jointly fit, executing \(L\) may remove the
fallback and the inequality \(c_L<p_LC_H\) is inapplicable.

Cheap failure can also change the continuation state. Let
\(q_0=p_H(s_0)>0\) be the success probability of direct escalation and
\[
  q_1=\mathbb E[p_H(S_1)\mid L\text{ fails}]
\]
the success probability of the same continuation after the
failure-conditioned successor state produced by \(L\). Across repeated
independent tasks, cheap-first improves amortized expected scalar cost per
accepted proof exactly when
\[
  \frac{c_L+(1-p_L)C_H}
       {p_L+(1-p_L)q_1}
  <
  \frac{C_H}{q_0},
\]
or equivalently,
\[
  c_L
  <
  \frac{C_H}{q_0}
  \left[p_L+(1-p_L)(q_1-q_0)\right].
\]
The two terms on the right separate immediate cheap success from the residual
effect of verifier-grounded state change on a fixed continuation policy. The
quantity \(q_1-q_0\) may be negative: a cheap action can consume useful context
or steer search toward a worse state.

\subsection{Verifier-Grounded Residual Progress}

The exact continuation difference above is unavailable online. We approximate
it using normalized numerical features \(\phi(W)\in\mathbb R^d\) derived from
the checker and workspace, and a potential
\[
  \Phi_{\mathbf w}(W)=\mathbf w^\top\phi(W).
\]
Features include accepted obligation mass, verified-prefix information, and
the number of newly ready dependents. Categorical goal fingerprints and
diagnostic transitions are action features or repetition penalties rather than
coordinates subtracted numerically. The transition residual is
\[
  \Delta_\Phi(W,a,W')
  =\Phi_{\mathbf w}(W')-\Phi_{\mathbf w}(W).
\]

\begin{proposition}[Telescoping residual]
For any trajectory \(W_0,\ldots,W_T\),
\[
  \sum_{t=0}^{T-1}\Delta_\Phi(W_t,a_t,W_{t+1})
  =\Phi_{\mathbf w}(W_T)-\Phi_{\mathbf w}(W_0).
\]
\end{proposition}

The identity follows by cancellation. If \(\Phi(W)\) is only the terminal
acceptance indicator, the current state is nonterminal, and a one-step action
accepts with probability \(p\), then
\(\mathbb E[\Delta_\Phi]/c=p/c\). Richer features give a denser but empirical
surrogate; they are not assumed to be monotone distance to a proof. To avoid
double counting in the controller, terminal acceptance, downstream unlock,
and nonterminal residual progress are logged as distinct terms, and their
weights or ranks are frozen without using test-task outcomes.
