\section{Budget-Aware Structured Control}

\subsection{Three Independent Control Levers}

At a decision point the controller receives
\[
  s_t=(W_t,\mathcal H_t,\mathbf b_t),
\]
where \(W_t\) is the workspace, \(\mathcal H_t\) contains only cost and
outcome statistics available before the next result, and \(\mathbf b_t\) is the
remaining resource vector. We separate three decisions:

\begin{enumerate}
    \item the \emph{action mask} determines which typed computations may be
    proposed in the current experiment;
    \item the \emph{selector} chooses a feasible branch--action pair; and
    \item the \emph{router} chooses the model tier used to propose or execute
    that computation.
\end{enumerate}

This separation is both an implementation boundary and an experimental one. A
larger action mask is not evidence for a better selector, and access to a
strong model is not evidence for adaptive routing.

\subsection{Cost Measurement and Vector Admission}

Every provider request, reported token quantity, tool call, checker call,
pricing event, and charge is appended to a common ledger. A measurement is
marked observed, estimated, unavailable, or unbounded; unavailable quantities
are never converted to zero. Action-cost estimates
\(\hat{\mathbf c}(s,n,a)\) come either from frozen static priors or from a
validation-only history snapshot bucketed by action kind and model tier. The
snapshot, estimator version, sample count, fallback reason, and realized
estimate error are retained in the trace.

Hard feasibility remains vector-valued. Before an action is executed, the
controller forms
\[
  \mathcal A_{\mathrm{feas}}(s,\mathbf b)
  =\left\{(n,a):
  \overline c_j(s,n,a)
  \le b_j-b_j^{\mathrm{reserve}}
  \text{ for every enforced resource }j\right\}.
\]
The final-check reserve is never available to ordinary branch actions.
Dimensions that cannot be compared are recorded explicitly, and the
preregistered policy determines whether an unknown estimate is rejected or
admitted under another hard guardrail. Proposal-generation cost is charged
separately from action-execution cost, so a structural action with no immediate
checker call is not incorrectly treated as free.

Per-obligation soft envelopes distribute checks and model requests across the
dependency graph. A high-unlock obligation may borrow from its envelope when
the global vector still admits the action; every such borrow is logged. These
envelopes guide allocation but never override a global hard limit.

\subsection{Verifier-Grounded Marginal Value}

For state \(s\), branch \(n\), and action \(a\), the value decomposition uses:

\begin{itemize}
    \item \(\hat p_{\mathrm{acc}}\), immediate checker-and-safety acceptance;
    \item \(\hat\delta_{\mathrm{fail}}\), nonterminal residual progress
    conditional on the action not accepting immediately;
    \item \(U_{\mathrm{unlock}}\), dependents made ready if the target is
    accepted;
    \item \(\hat I\), the value of discriminating among explicit competing
    failure hypotheses; and
    \item \(\hat R\), repetition and mutation-scope risk, including unchanged
    goal fingerprints.
\end{itemize}

One convenient cardinal template is
\[
  \widehat G(s,n,a)=
  \alpha\hat p_{\mathrm{acc}}
  +\beta(1-\hat p_{\mathrm{acc}})\hat\delta_{\mathrm{fail}}
  +\eta\hat p_{\mathrm{acc}}U_{\mathrm{unlock}}
  +\kappa\hat I
  -\rho\hat R.
\]
The conditional factors prevent immediate acceptance, failure progress, and
acceptance-dependent unlocks from being counted as the same event. When a
scalar comparison is required, a validation-frozen nonnegative shadow-price
vector gives
\[
  Q(s,n,a)=
  \frac{\widehat G(s,n,a)}
       {\epsilon+\boldsymbol\lambda^\top
        \hat{\mathbf c}(s,n,a)}.
\]
This expression is a policy template, not a theorem that the terms form a
calibrated utility or that \(Q\) is globally optimal.

The auditable heuristic instantiation uses deterministic, predeclared ranks
when a probability or cardinal value is unavailable. Readiness and hard
admission are gates; estimated execution cost, stalled streak, goal change,
unlock count, hypothesis-discriminating tests, and stable identifiers form the
rank and tie breakers. The value-only control removes estimated cost from this
rank. The value-per-cost control uses the same state signals and action mask,
and the remaining-budget control additionally rejects actions whose estimated
cost cannot fit after reserve. Exact keys, weights, cost snapshots, and fallback
behavior are serialized with every formal run and are frozen before test
evaluation.

Unlike a positive power transform of a deterministic score, vector admission
can change the selected action: it removes computations that would crowd out a
binding resource or final assembly. We therefore make no claim that a monotone
rescaling of \(Q\) alone makes a deterministic policy budget-conditioned.

\subsection{Diagnostic-Conditioned Actions}

Lean diagnostics alter the conditional value of computation. Parser and local
type errors often favor repair; unknown identifiers favor retrieval or a
capability test; a repeated unchanged goal lowers the rank of another
equivalent implementation attempt; and a sound-looking argument with
persistent implementation failures may favor a representation change. These
rules initialize heuristic features and action applicability. They do not turn
diagnostics into infallible mathematical diagnoses.

Competing failure hypotheses make information-seeking actions explicit. A
signature test may distinguish a missing declaration from an incorrect
argument and thereby avoid an expensive wrong route. We call this a
hypothesis-discrimination feature unless a probabilistic information-gain
model is actually fitted and calibrated. Infrastructure failures are handled
by deterministic runtime rules and are never presented as mathematical
hypotheses.

\subsection{Cheap/Strong Model Routing}

Both model tiers implement the same proof-agent protocol and share the same
workspace and ledger. The adaptive router defaults to the cheap tier and may
request the strong tier for structural actions, repeated failures on the same
goal fingerprint, repeated proposal-validation failures, a stalled repair
streak, or a high-unlock state with poor frozen cheap-tier success. Escalation
is granted only if the strong-tier cost estimate fits the remaining vector and
reserve. Thresholds and tier-specific cost estimates are selected on validation
data and frozen.

The cheap-first proposition in Section~3 motivates this router but does not
prove it optimal: real actions change state and their probabilities are only
estimated. The routing ablation therefore compares adaptive routing with a
validation-frozen escalation schedule that has the \emph{same} cheap/strong
model portfolio, tier caps, action mask, selector, and total resource vector.
Cheap-only and strong-only runs are useful boundary policies, but are not
interpreted as isolated routing effects.

\subsection{Controller Loop}

The structured loop is:

\begin{enumerate}
    \item initialize the workspace and seed all ready branches;
    \item reserve resources required for final assembly;
    \item enumerate actions permitted by the experimental mask;
    \item estimate costs, filter stale or vector-infeasible actions, and record
    the full rejected choice set;
    \item rank the remaining branch--action pairs and route the selected
    computation to an admitted model tier;
    \item execute it, charge realized cost, and deterministically reduce the
    result into a successor workspace;
    \item if a compatible solution exists, assemble and recheck it; otherwise
    continue until no useful feasible action remains.
\end{enumerate}

On termination, the system returns an accepted proof or a structured partial
result containing accepted helpers, blocked or open obligations, preserved
alternatives, raw checker observations, exact resource use, measurement
coverage, and the reason no further action was selected.
