\section{Results}

This section is a fillable reporting shell. It contains no synthetic
measurements; every final value must be regenerated from per-task results and
traces under the frozen configuration hash.

\subsection{Main Fixed-Budget Comparison}

\begin{table}[ht]
\centering
\small
\begin{tabular}{lrrrrr}
\toprule
Policy & Accepted/\(N\) & \(\Delta\) pp & Tokens/task & Checks/task & Std. \$/task \\
\midrule
Independent sampling & -- & -- & -- & -- & -- \\
Direct strong & -- & -- & -- & -- & -- \\
Minimal refinement & -- & -- & -- & -- & -- \\
Fixed escalation & -- & -- & -- & -- & -- \\
\system & -- & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\caption{Matched-budget results for
\todo{benchmark, model, and exact resource vector}. Costs are unconditional
means over attempted tasks. Add paired 95\% intervals, actual billed cost with
coverage, and the reference policy for \(\Delta\).}
\label{tab:main}
\end{table}

\todo{Lead with absolute accepted counts and the percentage-point difference,
then give the paired interval and unconditional cost difference. State whether
the result supports higher Pass@budget, noninferior success at lower cost, or
neither. Do not lead with a relative percentage improvement.}

Each benchmark--model cell receives its own panel. A secondary summary reports
selected, completed, accepted, proof-failure, infrastructure-failure, retried,
and missing counts, plus the fraction of runs that hit each resource cap.

\subsection{Solve--Cost Frontier Across Budgets}

\todo{Plot accepted rate against standardized dollars, logical tokens, checker
calls, and the preregistered nested budget profiles. Show confidence bands and
actual mean expenditure at every cap. Report if policy rankings change between
low-, medium-, and high-budget regimes; do not project incomparable resource
vectors onto one undeclared scalar curve.}

For noninferiority claims, report the accepted-rate difference and margin next
to the ratio of unconditional resource use. Successful-run cost and amortized
cost per accepted proof appear as separately labeled diagnostics.

\subsection{Controlled Mechanism Effects}

\begin{table}[ht]
\centering
\small
\begin{tabular}{llrrrr}
\toprule
Factor & Matched contrast & \(\Delta\) pp & Token ratio & Cost ratio & NI? \\
\midrule
Action space & \(A_{10}-A_{00}\) & -- & -- & -- & -- \\
Action space & \(A_{11}-A_{01}\) & -- & -- & -- & -- \\
Cost-aware selector & \(A_{01}-A_{00}\) & -- & -- & -- & -- \\
Cost-aware selector & \(A_{11}-A_{10}\) & -- & -- & -- & -- \\
Cost normalization & \(S_{\mathrm{ratio}}-A_{10}\) & -- & -- & -- & -- \\
Remaining admission & \(A_{11}-S_{\mathrm{ratio}}\) & -- & -- & -- & -- \\
Adaptive routing & \system{}-\(R_{\mathrm{fixed}}\) & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\caption{Mechanism-isolating contrasts under one structured harness. Ratios
use unconditional per-task resource consumption; report paired intervals and
the action-space--selector interaction.}
\label{tab:mechanisms}
\end{table}

\todo{Interpret only the factor changed by each contrast. If the
action-space--selector interaction is material, report the two conditional
effects rather than presenting one additive decomposition. If \system{} beats
minimal refinement but not the matched selector or routing control, revise the
headline claim accordingly.}

\subsection{Action Evidence and Task Regimes}

\begin{table}[ht]
\centering
\small
\begin{tabular}{lrrrrrr}
\toprule
Action & Eligible & Proposed & Executed & Progress & Accepted & Exclusive \\
\midrule
Capability test & -- & -- & -- & -- & -- & -- \\
Decompose & -- & -- & -- & -- & -- & -- \\
Refine argument & -- & -- & -- & -- & -- & -- \\
Change representation & -- & -- & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\caption{Evidence for richer-only actions. The Exclusive column counts paired
tasks solved by the expanded mask and not by its restricted control. Define
positive progress from the frozen verifier-grounded transition rule.}
\label{tab:actions}
\end{table}

\todo{Stratify the end-to-end comparison by preregistered task properties,
including initial scaffold size, observed error class, obligation count, and
dependency depth. Estimate the crossover at which structured control recovers
its prompt/control overhead, and report simple tasks where minimal refinement
is strictly cheaper. Treat post-run difficulty strata as exploratory.}

\subsection{Decision and Failure Analysis}

\todo{Report probability calibration for explicit success estimates, ranking
lift for non-probabilistic priorities, cost-estimate coverage and error,
escalation timing, repeated-state checks, and checks outside final artifact
ancestry. Include representative paired traces: one allocation win, one case
where preserving an alternative matters, and one failure where the controller
spends budget poorly. Separate infrastructure faults from mathematical proof
failures.}
