\section{Limitations and Broader Considerations}

Our exact results are deliberately local. The \(p_i/c_i\) theorem fixes the
candidate set and minimizes expected expenditure until success or exhaustion;
under hard truncation, selecting that set is a knapsack problem and greedy
ratio selection need not maximize accepted rate. The cheap-prefix criterion
applies only when the prefix and fallback are jointly feasible. With multiple
resources, ratio ordering also requires a prespecified scalarization, while
hard feasibility remains component-wise. Real proof actions update beliefs,
change future action sets, and have dependent outcomes, so the controller's
richer score is a policy to evaluate rather than a proof of global optimality.

Verifier-grounded residual progress is a surrogate for continuation value, not
distance to a proof. A candidate with fewer goals may have replaced an easy
goal with a harder one; a decomposition can inflate the number of accepted
helpers without bringing the root closer to assembly; and unlock, progress,
and immediate success features can overlap if they are not defined carefully.
Weights, ranks, normalization, and cost histories must be frozen on validation
data and checked for calibration or ranking lift across domains and models.

Multi-resource comparison has no unique scalar efficiency measure. Tokens,
checker calls, latency, and dollars can bind different policies, and provider
pricing or token reporting can be incomplete. We therefore report the full
budget and realized cost vectors, standardized and billed monetary cost, and
measurement coverage. Robustness to alternative shadow prices does not replace
component-wise feasibility.

Structured search has overhead. It uses longer projections, executes control
logic, and may spend early budget constructing obligations that a direct model
could solve immediately. Minimal refinement remains a first-class end-to-end
baseline, and we report task-stratified crossover behavior rather than claiming
that structured execution should be universal. Conversely, minimal versus
structured execution is not a clean representation ablation because their
contexts and control loops differ; mechanism attribution relies on matched
structured arms.

The controlled factors can interact. An expanded action space changes the
states visited by a selector, and adaptive routing may be valuable only for
structural actions. Marginal ablation effects are therefore conditional on the
held-fixed configuration and need not sum to the full-system effect. A second
public benchmark and frozen model/budget cells are required before a
cross-domain claim; local multi-obligation stress tests cannot substitute for
that confirmation.

The current system targets Lean~4. The proof-system adapter is narrow enough to
make another backend plausible, but obligation semantics, diagnostics, tactic
granularity, libraries, and checker cost differ across assistants. Results
should not be generalized to Isabelle, Rocq, or Metamath without new
experiments.

Finally, a checker-accepted proof certifies the formal statement, not the
faithfulness of an automatically formalized statement to the user's intent.
The primary experiments fix formal statements. Statement preservation and
anti-cheating checks prevent several mechanical shortcuts, but they do not
replace human review of high-stakes specifications.
